---
title: "Endangered Languages of the World"
subtitle: "INFO 526 - Summer 2026 - Final Project"
author: 
  - name: "Damon Kelson"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "An exploration of the features of endangered languages"
format:
   html:
    code-tools: true
    code-overflow: wrap
    embed-resources: true
editor: visual
execute:
  warning: false
  echo: false
---

## Abstract

Linguistic analysis reveals historical and cultural details that track the geographic distribution of human populations. Endangered languages represent the potential extinction of certain subsets of human culture containing unique myths and history. This project aims to analyze the global spread of endangered languages, evaluating both their geography and numerical features of their language family, to better understand what languages are most threatened.

All code below taken from presentation.qmd and adapted to index.qmd's visual format when needed, as the presentation was developed first. 

```{r data setup carried over from presentation slides and prep}
#| include: false

#pacman::p_load(tidyverse, kableExtra, sf, patchwork, rnaturalearth, rnaturalearthdata)

# I'm replacing the pacman load because the git website publishing is failing due to not loading via pacman
# Hopefully this explicit method fixes the publishing issue
library(tidyverse)
library(kableExtra)
library(forcats)
library(sf)
library(ggrepel)
library(patchwork)
library(rnaturalearth)
library(rnaturalearthdata)

# Reading in the requisite data
languages <- read_csv("data/languages.csv", show_col_types = FALSE)
status    <- read_csv("data/endangered_status.csv", show_col_types = FALSE)
families  <- read_csv("data/families.csv", show_col_types = FALSE)

# Ordering status labels
status_order <- status |>
  distinct(status_code, status_label) |>
  arrange(status_code) |>
  pull(status_label)

# Unclear groups are groups of languages that do not easily fit in this analysis
# All together, this is about 10% of the data!
unclear_groups <- c("Bookkeeping", "Unclassifiable", "Pidgin",
                    "Speech Register", "Sign Language", "Unattested",
                    "Artificial Language", "Mixed Language")

# Family size determination for later use
family_size <- languages |>
  filter(!is.na(family_id)) |>
  count(family_id, name = "family_size")

# Defined colors for common usage across some later plots
risk_colors <- c(
  "not endangered" = "#2A6F97",
  "at risk"        = "#E9A23B",
  "extinct"        = "#B23A48"
)

# The base level dataset, with joins
lang_base <- languages |>
  left_join(status, by = "id") |>
  left_join(families, by = c("family_id" = "id")) |>
  left_join(family_size, by = "family_id") |>
  mutate(
    status_label = factor(status_label, levels = status_order, ordered = TRUE),
    # 3-category split: 1 | 2,3,4,5 | 6... NA kept around (~5%)
    risk_group = case_when(
      status_code == 1     ~ "not endangered",
      status_code %in% 2:5 ~ "at risk",
      status_code == 6     ~ "extinct"
    ),
    risk_group = factor(risk_group, levels = c("not endangered", "at risk", "extinct")),
    is_unclear_group = family %in% unclear_groups,
    family_bucket = case_when(
      is_isolate       ~ "isolate",
      is_unclear_group ~ "unclear (exclude)",
      family_size <= 10  ~ "small (2-10)",
      family_size <= 50  ~ "medium (11-50)",
      family_size > 50   ~ "large (50+)"
    ),
    family_bucket = factor(family_bucket, levels = c(
      "isolate", "small (2-10)", "medium (11-50)", "large (50+)", "unclear (exclude)"
    ))
  )

lang_geo <- lang_base |> filter(!is.na(latitude), !is.na(longitude))

# map with extinction dropped
map_data_excl_extinct <- lang_geo |>
  filter(risk_group %in% c("not endangered", "at risk")) |>
  mutate(at_risk_num = as.numeric(risk_group == "at risk"))

# per family risk severity breakdown
family_severity <- lang_base |>
  filter(!is.na(risk_group), family_bucket != "unclear (exclude)") |>
  count(family_bucket, risk_group) |>
  group_by(family_bucket) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  mutate(
    # Ease of controlling bar formatting
    family_bucket = fct_drop(family_bucket),
    label_size = if_else(risk_group == "extinct", 4, 3),
    label_face = if_else(risk_group == "extinct", "bold", "plain")
  )

```


## Introduction

This project utilizes the three TidyTuesday datasets from Glottolog, a comprehensive reference catalog of the world's languages. The core dataset covers over 8000 languages and their accompanying family, geographic coordinates, macroarea label, and country associations. The endangered_status.csv associates each language with a 1-6 scale, (1) being not endangered through shifting, threatened, moribund, nearly extinct, to full extinct (6). 

In my following analysis, I collapsed the endangered statuses down to 3 categories: not endangered (code 1), at risk (codes 2 through 4), and extinct (code 6). This was done to create a broad distinction between languages in the middle which are facing varying levels of threats to their survival, from languages which are doing fine to those that are no longer around (extinct languages of course being no longer "under threat"). An important note on what I've grouped as "unclear groups": these are languages such as sign languages, pidgins, artificial languages (e.g. esperanto, fantasy fiction, etc.) and so on which were either inappropriate for this analysis or were unfortunately not attested strongly enough in the data set to meaningfully analyze. Unfortunately "unclassifiable" and "unattested" DO contain interesting languages relevant to this analysis, but their completeness in the dataset was not adequate for the scope of this analysis or the time available.

## Question 1 - How does geography correlate with endangered language status?

### Intro:

For this question, I explored where endangered languages clustered geographically both by macroarea (continent/large geographical area) as well as regionally via their lat/long coordinates. Examining the proportion of threatened languages at both the macroarea and regional grid levels allows us to analyze endangerment patterns across multiple scales while also incorporating raw macroarea language counts for overall context. Forces threatening languages and cultures originate from many diverse causes far too complicated to explore in-depth for a project such as this. However, many notable threats tend to follow geographic patterns such as: national assimilation programs, minority political marginalization, and urbanization.

### Approach:

As I explored different visualization possibilities, first starting with the hexbin plots, I had difficulty reconciling the sparse but highly threatened nature of certain regions (e.g. many North American natives) with the extreme density of other threatened regions such as Papunesia (Papua New Guinea & parts of Oceania). My compromise was to create a choropleth for macroareas accompanied by a table so that I could show and discuss how much the density of endangered languages vary by region globally, thereby freeing up the hexbin plot from needing to address anything other than the percent of endangered languages in a given hexbin.

### Analysis:

```{r Q1 specific data prep}
# country polygons
world <- ne_countries(scale = "medium", returnclass = "sf")

# ECK4 equal-area pseudocyrlindrical projection
# initiated before grid cells to preserve equal area across latitudes
eck4 <- "+proj=eck4"

lang_sf    <- st_as_sf(map_data_excl_extinct, coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(eck4)
world_proj <- st_transform(world, eck4)

# projected CRS, cellsize in meters, hexagon shape
# cellsize tuning for hexplot
grid <- st_make_grid(world_proj, cellsize = 900000, square = FALSE) |>
  st_as_sf() |>
  mutate(cell_id = row_number())

# per cell % at risk, n_lang
cell_summary <- st_join(lang_sf, grid) |>
  st_drop_geometry() |>
  group_by(cell_id) |>
  summarize(pct_at_risk = mean(at_risk_num), n_lang = n(), .groups = "drop")

# drop empty cells, constrain representations of large differences
grid_plot <- grid |>
  left_join(cell_summary, by = "cell_id") |>
  filter(!is.na(n_lang))

# Creating macroarea representations
country_macro <- map_data_excl_extinct |>
  # Parsing lists of multiple macroarea associations and creating 1
  filter(!is.na(macroarea), !str_detect(macroarea, ";"), !is.na(countries)) |>
  separate_longer_delim(countries, delim = ";") |>
  count(countries, macroarea) |>
  group_by(countries) |>
  slice_max(n, n = 1, with_ties = FALSE) |>
  ungroup()

# Uruguay parses as "Eurasia" due to heavy presence of foreign languages
# This exists to explicitly correct the above for the only failure I can find
country_macro <- country_macro |>
  mutate(
    macroarea = if_else(countries == "UY", "South America", macroarea)
  )

macro_summary <- map_data_excl_extinct |>
  filter(!is.na(macroarea), !str_detect(macroarea, ";")) |>
  group_by(macroarea) |>
  summarize(pct_at_risk = mean(at_risk_num), n_lang = n(), .groups = "drop")

# Least error-prone join between rnaturalearth and my base dataset coords
country_risk <- world |>
  left_join(country_macro, by = c("iso_a2_eh" = "countries")) |>
  left_join(macro_summary, by = "macroarea")
```

:::: {layout="[[70,30]]"}

::: {}

```{r Q1 macroarea plot}

ggplot(country_risk) +
  geom_sf(aes(fill = pct_at_risk), color = "grey50", linewidth = 0.15) +
  scale_fill_gradient(low = "grey90", high = "#C0392B",
                      labels = scales::percent,
                      na.value = "white",
                      guide = guide_colorbar(
                        title.position = "top",
                        title.hjust = 0.5,
                        barwidth = unit(5, "cm")
                      )
  ) +
  labs(title = "Endangerment Risk by Geographic Macroarea",
       subtitle = "Regions colored by the share of at-risk languages in their macroarea (extinct excluded)",
       caption = "Glottolog 5.2.1 (via TidyTuesday, 2025-12-23)",
       fill = "% of languages at risk") +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank())
```

:::

::: {}

```{r Q1 macroarea table}
macro_summary |>
  arrange(desc(pct_at_risk)) |>
  transmute(Region = macroarea,
            `% at risk` = scales::percent(pct_at_risk, accuracy = 1),
            `N languages` = n_lang) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

:::

::::

```{r Q1 hexbin plot}
#| fig-width: 9
#| fig-height: 5
#| out-width: 100%
#| fig-align: center

ggplot() +
  geom_sf(data = world_proj, fill = "grey97", color = "grey90", linewidth = 0.15) +
  geom_sf(data = grid_plot, aes(fill = pct_at_risk), color = NA) +
  geom_sf(data = world_proj, fill = NA, color = "grey30", linewidth = 0.2) +
  scale_fill_gradient(low = "grey90",
                      high = "#C0392B",
                      labels = scales::percent,
                      guide = guide_colorbar(
                        title.position = "top",
                        title.hjust = 0.5,
                        barwidth = unit(5, "cm")
                      )) +
  coord_sf(crs = eck4) +
  labs(title = "Regional Patterns of Language Endangerment",
       subtitle = "Language risk summarized in equal-area hexagons",
       caption = "Glottolog 5.2.1 (via TidyTuesday, 2025-12-23)",
       fill = "% of languages at risk") +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        plot.title = element_text(size=18)
        )
```


### Discussion:

Both the macroarea and hexbin plots show a clear high-proportion of endangered languages across the Americas as well as Australia. However, Papunesia and eastern Eurasia do not convey an especially endangered-looking situation in the macroarea plot. In contrast, the hexbin plot more clearly demonstrates how many regions in central and eastern Siberia contain endangered languages, when separated from the more numerous and less proportionately endangered regions of Europe/Middle east/South Asia. The macroarea plot with its table excels relative to the hexbin approach by supplying context; as dire as the situation in Australia appears (99% at risk, 134 languages), Papunesia's absolute magnitude of linguistic diversity (65% at risk, 2041 languages!) defies understanding with any attempt at characterization that's not highly regional.

When the two plots are taken together it paints a picture of high threat regions in North America, South America, Australia, New Guinea, Oceania, and eastern Eurasia. Over the past 5 centuries these regions have been at the forefront of colonial activity, with local hunter-gatherer populations being especially vulnerable to subjugation/assimilation/lingua franca adoption pressure. As an interesting contrast, Africa is one of the least threatened macroareas for linguistic diversity. Political fragmentation and the highly local/regional nature of governance throughout the continent seems to be a likely contributor to this resiliency.

## Question 2 - How does language family size/isolate status vary for endangered and extinct languages?

### Intro:

Language isolates are languages which are not known to be related to any other language. They represent highly vulnerable branches of human culture, by definition they are the last known surviving member of their linguistic branch. Initially this question was focused primarily on the numerical proportions of extinct languages by family size. Exploring this question began without any solid geographical or regional consideration, but very early on I realized this question was intimately tied to two regions: South America and Papua New Guinea. Depending on how you define the regions exactly, these two together constitute over 60% of all known isolates.

### Approach:

The analysis is largely centered on the "is_isolate" flag from the dataset, but the second visualization pulls heavily from the previous geographical preparation in order to highlight the two regions of interest. A 100% stacked bar chart was a natural fit for comparisons among family groups by their relative proportions of not endangered/at risk/extinct, and I wanted them to "flow downward" in a sense to highlight the growing proportion of "extinct" as the family became smaller. The second visual was meant to accompany the narrative the bar chart builds, which is that very few isolates are "not endangered", only 3% in-fact. With 66% of all isolates in the world contained in these two regions, the narrative of "high risk" for isolates provides powerful context for how especially threatened these two regions are.

### Analysis:

```{r Q2 data prep}
iso_geo <- lang_base |> filter(is_isolate, !is.na(latitude), !is.na(longitude))
iso_sf  <- st_as_sf(iso_geo, coords = c("longitude", "latitude"), crs = 4326, remove = FALSE)

# map crops/isolate regions, also used in calculations
sa_box <- list(lon = c(-100, -20), lat = c(-35, 22))
oc_box <- list(lon = c(110, 175), lat = c(-45, 5))

iso_region <- iso_sf |>
  mutate(
    region_focus = case_when(
      between(latitude, sa_box$lat[1], sa_box$lat[2]) & between(longitude, sa_box$lon[1], sa_box$lon[2]) ~ "Northern South America",
      between(latitude, oc_box$lat[1], oc_box$lat[2]) & between(longitude, oc_box$lon[1], oc_box$lon[2]) ~ "Australia, New Guinea & Oceania",
      TRUE ~ "elsewhere"
    )
  ) |>
  st_drop_geometry()

focus_pct <- mean(iso_region$region_focus != "elsewhere")
```

```{r Q2 stacked bars}
#| fig-width: 9
#| fig-height: 5
#| out-width: 90%
#| fig-align: center


ggplot(family_severity, aes(x = family_bucket, y = n, fill = risk_group)) +
  geom_col(position = "fill", width = 0.7, color = "black", linewidth = 0.3) +
  geom_text(
    aes(label = scales::percent(pct, accuracy = 1), size = label_size, fontface = label_face,
        group = risk_group),
    position = position_fill(vjust = 0.5), color = "white"
  ) +
  scale_size_identity() +
  scale_fill_manual(values = risk_colors) +
  guides(fill = guide_legend(reverse = TRUE)) +
  coord_flip() +
  labs(x = NULL, y = NULL, fill = NULL,
       title = "Language Endangerment by Family Size",
       caption = "Glottolog 5.2.1 (via TidyTuesday, 2025-12-23)") +
  theme_minimal() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        panel.grid.major.x = element_blank(), panel.grid.minor.x = element_blank(),
        legend.position = "top")
```

```{r Q2 two isolate region plots}
#| fig-width: 9
#| fig-height: 5
#| out-width: 90%
#| fig-align: center


p_map_base <- ggplot() +
  geom_sf(data = world, fill = "grey95", color = "grey90", linewidth = 0.2) +
  geom_point(data = iso_geo, aes(longitude, latitude, color = risk_group), alpha = 0.7,
             size = 1.6, position = position_jitter(width = 0.2, height = 0.2, seed = 1)) +
  scale_color_manual(values = risk_colors) +
  theme_void() +
  theme(legend.position = "none")

p_sa <- p_map_base +
  coord_sf(xlim = sa_box$lon, ylim = sa_box$lat, expand = FALSE) +
  labs(title = "Northern South America", color = "Endangerment Status") +
  theme(
  legend.position = c(0.97, 0.97),
  legend.justification = c(1, 1),
  legend.margin = margin(3, 4, 3, 4),
  legend.key.size = unit(0.35, "cm"),
  legend.title = element_text(size = 8, face = "bold"),
  legend.text = element_text(size = 7),
  legend.background = element_rect(fill = "white", color = "black", linewidth = 0.6),
  plot.margin = margin(t = 5.5, r = 8, b = 5.5, l = 5.5),
  plot.title = element_text(hjust = 0.5, face = "bold", size = 14, margin = margin(b = 15))
) +
  guides(color = guide_legend(override.aes = list(size = 1.8)))

p_oc <- p_map_base +
  coord_sf(xlim = oc_box$lon, ylim = oc_box$lat, expand = FALSE) +
  labs(title = "Australia, New Guinea & Oceania") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14, margin = margin(b = 15)))

(p_sa | p_oc) +
  plot_annotation(
    title = paste0(scales::percent(focus_pct, accuracy = 1),
                   " of Language Isolates are Concentrated in Two Regions"),
    caption = "Glottolog 5.2.1 (via TidyTuesday, 2025-12-23)",
    theme = theme(plot.title = element_text(size = 18, face = "bold", hjust = 0.5,
                                             margin = margin(b = 20)))
  )
```


### Discussion:

The trend towards a higher proportion of extinct languages as the family size shrinks is far stronger than I had initially guessed, and unraveling why this is the case would be even more daunting than attempting to explain the geographical narrative! There are competing ways to frame the situation: is an isolate, by nature of being part of a language family so diminished that it's the only known member, simply the manifestation of us glimpsing the last stage of this trend? Do members of larger language families such as Indo-European or Niger-Congo provide a subtle support network for other intra-familial languages which bolsters their survival? Well beyond the scope of this project! Nevertheless there appears to be a vulnerability to being the last known language of a family, and I'd expect the causes to be quite dynamic and varied.

One trend that is clear is the regional distribution of language isolates. With 66% of all isolates in the cropped regions of South America and the Papunesia macroarea, they tell very similar but subtly distinct stories. These regions have been extremely isolated historically; both are defined by dramatically isolated valleys and inaccessibly dense jungle. The geographical nature of these regions maximizes isolation for pre-modern societies, so it's no wonder so many isolates hold on in obscure locales. However, the visualization also demonstrates that these regions occupy two different positions in endangerment. South America shows a higher proportion of extinct isolates than New Guinea, almost certainly the result of centuries of Spanish colonial expansion and stronger political centralization. New Guinea, by contrast, has retained a far higher proportion of non-extinct isolates, likely owing to its less central position in colonial projects and the persistent weak centralized government.

