---
title: "Endangered Languages of the World"
subtitle: "Patterns in threatened languages"
author: 
  - name: "Damon Kelson"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "An exploration of the features of endangered languages"
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false
library(tidyverse)
```

## Dataset

```{r}
#| label: load-dataset
#| message: false

endangered_status <- read_csv("data/endangered_status.csv")
families <- read_csv("data/families.csv")
languages <- read_csv("data/languages.csv")

# TidyTuesday readme page:
# https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-12-23/readme.md
```

TidyTuesday's "The Languages of the World" dataset comes from Glottolog 5.2.1's collaborative catalog of linguistic data. A quick glimpse at our 3 csv's:

**Endangered status**

```{r}
glimpse(endangered_status)
unique(endangered_status$status_code)
```

**Families**

```{r}
glimpse(families)
```

**Languages**

```{r}
glimpse(languages)
```

From our glimpses we can see that "endangered_status.csv" contains the necessary lookup values for how threatened a given language is, "families.csv" contains the language family grouping labels, and "languages.csv" contains the bulk of data covering language name, geographic extent, whether it is an isolate, etc.

I've always found linguistics fascinating and it's a reoccurring topic I chose to study in my free time. Not only are languages themselves interesting, but linguistic analysis enables us to discern an unbelievable range of subtle details about the history and culture of those who spoke and/or wrote it. Endangered languages represent the potential extinction of certain subsets of human culture containing unique myths and history, and so understanding them may help prevent their decline or at least preserve some knowledge before they're lost.

## Questions

Question 1: How does geography correlate with endangered language status?

Question 2: What patterns exist in language family/isolate status for endangered languages?

## Analysis plan

**Question 1:**

- Variables: macroarea, latitude, longitude, language name, endangered status.
- Created variables: Potentially a binary between "endangered" and "not endangered" to visualize those languages in some significant level of danger from those that aren't in significant danger (or are extinct already).
- Usage of a world map for the plot is a natural fit here, perhaps using a heat map of endangered language prevalence to examine their distribution.

**Question 2:**

- Variables: language name, family, endangered status, isolate status.
- Created variables: Some aspect of the family (such as the number of members or even something representing total geographic extent).
- Analysis here will likely involve exploring whether or not language isolates are more likely to be endangered and whether languages in small families are more likely to be endangered.

| Task                                | Status | Due       | Summary |
|-------------------------------------|--------|-----------|---------|
| Explore possible visualizations     |        | 6/14/2026 |         |
| Create chosen analysis and visuals  |        | 6/21/2026 |         |
| Further refinement and presentation |        | 6/28/2026 |         |
